Transactional Memory Concurrency Verification with Landslide

### 2018-03-12 htm/paper.tex
txn structure deisgn section and write about tso/pso

\usepackage{yfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{url}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{harmony}
\usepackage{marvosym}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[charter]{mathdesign}
\usepackage{natbib}
\usepackage[mathcal]{euscript}
\usepackage[linesnumbered,ruled]{algorithm2e}

@@

% TODO

@@

{\bf Stateless model checking.}
Model checking (MC) \cite{verisoft} is a testing technique for systematically executing and verifying
the possible thread interleavings of a concurrent program.
The main research challenge is to cope with exponential explosion of the state space,
which is $O(n^k)$ for a program with $n$ operations and $k$ threads.
Some {\em stateful} MCs explicitly store and compare all visited states of the program being tested \cite{spin},
which both keeps track of test coverage
and allows identifying identical states to avoid testing redundant interleavings.
%
By contrast, {\em stateless} MC (henceforth abbreviated simply as MC)
stores only the current sequence of execution events to avoid a prohibitive memory footprint.
Reduction algorithms \cite{dpor,optimal-dpor,satcheck,mcr,mcr,tsopso}
can then analyze the memory accesses in that sequence to identify
interleavings observationally-equivalent under Mazurkiewicz trace theory \cite{mazurkiewicz} and hence safe to skip.
The resulting state spaces are still exponentially-sized,
but only in the number of conflicting operations rather than all operations.
Of these, Landslide uses Dynamic Partial Order Reduction \cite{dpor} to prune its state spaces.

@@

\section{Design}

Let $I = TN_1@L_1, TN_2@L_2, ... TN_n@L_n$,
with $N_i$ a thread ID and $L_i$ a code line number,
denote the execution sequence of a program as it runs according to the specified thread interleaving.%
\footnote{This serialization of concurrent execution is told from the perspective of main memory
and hence assumes sequential consistency.
For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.}

{\bf Example.}
Consider again the program in Figure~\ref{fig:htm-example}.
Note that the C-style {\tt x++} operations, when compiled into assembly,
results in multiple memory accesses which can be interleaved with other threads.
\[
\begin{tabular}{ll}
	$2a$ & \texttt{temp <- x;} \\
	$2b$ & \texttt{temp <- temp + 1;} \\
	$2c$ & \texttt{x <- temp;} \\
\end{tabular}
\]

@@

\newcommand\ti{\ensuremath{\hilight{lavender}{\mathbf{T1}}}}
\newcommand\tj{\ensuremath{\hilight{seafoam}{\mathbf{T2}}}}
\newcommand\tk{\ensuremath{\hilight{salmon}{\mathbf{T3}}}}

If these instructions from the {\tt x++} in the transaction are preempted,
with another thread's access to {\tt x} interleaved in between,
the transaction will abort.
So, the interleaving
\[
	\ti@1, \ti@2a, \ti@2b, \tj@1, \tj@2, \tj@3, \ti@2c, \ti@3
\]
or, henceforth abbreviated for clarity:
\[
	\ti@1-2b, \tj@1-3, \ti@2c-3
\]
is not possible; rather, \ti will fall into the backup path:
\[
	\ti@1-2b, \tj@1-3, \ti@4-7
\]
However, the {\tt x++} operation from the failure path (correspondingly $6a$, $6b$, $6c$)
can be thusly separated with conflicting accesses interleaved in between,
since the mutex only protects the failure path against other failure paths,
but not against the transaction itself.
So (assuming {\tt x} is intended to be a precise counter rather than a sloppy one),
%losing one of the increments to which constitutes a bug),
we observe a bug in the following interleaving.%
\footnote{Note also that this bug requires either at least 3 threads or at least 2 iterations between 2 threads to expose;
this highlights MC's dependence on its test cases to produce meaningful state spaces in the first place.}
\[
	\ti@1-2b, \tj@1-3, \ti@4-6b, \tk@1-3, \ti@6c-7
\]
Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
to exclude this family of interleavings,
showing that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
%further motivating our research.

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		  & \texttt{prevent\_transactions = false;} \\
		\\
		0 & \texttt{while (prevent\_transactions) continue;} \\
		1 & \texttt{if ((status = \_xbegin()) == SUCCESS) \{} \\
		2 & \texttt{~~~~if (prevent\_transactions)} \\
		3 & \texttt{~~~~~~~~\_xabort();} \\
		4 & \texttt{~~~~x++;} \\
		5 & \texttt{~~~~\_xend();} \\
		6 & \texttt{\} else \{} \\
		7 & \texttt{~~~~mutex\_lock(\&m);} \\
		8 & \texttt{~~~~prevent\_transactions = true;} \\
		9 & \texttt{~~~~x++;} \\
		A & \texttt{~~~~prevent\_transactions = false;} \\
		B & \texttt{~~~~mutex\_unlock(\&m);} \\
		C & \texttt{\}} \\
		%\texttt{~~~~\}} \\
		%\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{Variant of the program in Figure~\ref{fig:htm-example},
		with additional synchronization to protect the failure path from the transactional path.
		(The optional line 0 serves to prevent a cascade of failure paths
		for the sake of performance
		by allowing threads to wait until transacting is safe again.)}
	\label{fig:htm-fixed}
\end{figure}

{\bf Modeling transaction failure.}

@@

\section{Limitations}

\section{Future Work}
\label{sec:warpzone}

>>>>

\usepackage{yfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{url}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{harmony}
\usepackage{marvosym}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[charter]{mathdesign}
\usepackage{natbib}
\usepackage{xspace}
\usepackage[mathcal]{euscript}
\usepackage[linesnumbered,ruled]{algorithm2e}

@@

@@

{\bf Stateless model checking.}
Model checking (MC) \cite{verisoft} is a testing technique for systematically executing and verifying
the possible thread interleavings of a concurrent program.
The main research challenge is to cope with exponential explosion of the state space,
which is $O(n^k)$ for a program with $n$ operations and $k$ threads.
Some {\em stateful} MCs explicitly store and compare all visited states of the program being tested \cite{spin},
which both keeps track of test coverage
and allows identifying identical states to avoid testing redundant interleavings.
%
By contrast, {\em stateless} MC (henceforth abbreviated simply as MC)
stores only the current sequence of execution events to avoid a prohibitive memory footprint.
Reduction algorithms \cite{dpor,optimal-dpor,satcheck,mcr,mcr,tsopso}
can then analyze the memory accesses in that sequence to identify
interleavings observationally-equivalent under Mazurkiewicz trace theory \cite{mazurkiewicz} and hence safe to skip.
The resulting state spaces are still exponentially-sized,
but only in the number of conflicting operations rather than all operations.
Of these, Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to prune its state spaces.

@@

\section{Design}
\label{sec:design}

This section presents our formalization of transactional memory in Landslide's framework of thread concurrency.
We make two major simplifications:
simulating transaction aborts as failure injections,
and treating transaction atomicity as a global mutex during data-race analysis;
and provide corresponding equivalence proofs.

{\bf Notation.} Let $I = TN_1@L_1, TN_2@L_2, ... TN_n@L_n$,
with $N_i$ a thread ID and $L_i$ a code line number,
denote the execution sequence of a program as it runs according to the specified thread interleaving.%
\footnote{This serialization of concurrent execution is told from the perspective of main memory
and hence assumes sequential consistency.
For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.}

\subsection{Example}

Consider again the program in Figure~\ref{fig:htm-example}.
Note that the C-style {\tt x++} operations, when compiled into assembly,
results in multiple memory accesses which can be interleaved with other threads.
\[
\begin{tabular}{ll}
	$2a$ & \texttt{temp <- x;} \\
	$2b$ & \texttt{temp <- temp + 1;} \\
	$2c$ & \texttt{x <- temp;} \\
\end{tabular}
\]

@@

\newcommand\ti{\ensuremath{\hilight{lavender}{\mathbf{T1}}}\xspace}
\newcommand\tj{\ensuremath{\hilight{seafoam}{\mathbf{T2}}}\xspace}
\newcommand\tk{\ensuremath{\hilight{salmon}{\mathbf{T3}}}\xspace}

If these instructions from the {\tt x++} in the transaction are preempted,
with another thread's access to {\tt x} interleaved in between,
the transaction will abort.
So, the interleaving
\[
	\ti@1, \ti@2a, \ti@2b, \tj@1, \tj@2, \tj@3, \ti@2c, \ti@3
\]
or, henceforth abbreviated for clarity:
\[
	\ti@1-2b, \tj@1-3, \ti@2c-3
\]
is not possible; rather, \ti will fall into the backup path:
\[
	\ti@1-2b, \tj@1-3, \ti@4-7
\]
However, the {\tt x++} operation from the failure path (correspondingly $6a$, $6b$, $6c$)
{\em can} be thusly separated with conflicting accesses interleaved in between,
since the mutex only protects the failure path against other failure paths,
but not against the transaction itself.
So (assuming {\tt x} is intended to be a precise counter rather than a sloppy one),
%losing one of the increments to which constitutes a bug),
we observe a bug in the following interleaving.%
\footnote{Note also that this bug requires either at least 3 threads or at least 2 iterations between 2 threads to expose;
this highlights MC's dependence on its test cases to produce meaningful state spaces in the first place.}
\[
	\ti@1-2b, \tj@1-3, \ti@4-6b, \tk@1-3, \ti@6c-7
\]
Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
to exclude this family of interleavings,
showing that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
%further motivating our research.

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		  & \texttt{prevent\_transactions = false;} \\
		\\
		0 & \texttt{while (prevent\_transactions) continue;} \\
		1 & \texttt{if ((status = \_xbegin()) == SUCCESS) \{} \\
		2 & \texttt{~~~~if (prevent\_transactions)} \\
		3 & \texttt{~~~~~~~~\_xabort();} \\
		4 & \texttt{~~~~x++;} \\
		5 & \texttt{~~~~\_xend();} \\
		6 & \texttt{\} else \{} \\
		7 & \texttt{~~~~mutex\_lock(\&m);} \\
		8 & \texttt{~~~~prevent\_transactions = true;} \\
		9 & \texttt{~~~~x++;} \\
		A & \texttt{~~~~prevent\_transactions = false;} \\
		B & \texttt{~~~~mutex\_unlock(\&m);} \\
		C & \texttt{\}} \\
		%\texttt{~~~~\}} \\
		%\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{Variant of the program in Figure~\ref{fig:htm-example},
		with additional synchronization to protect the failure path from the transactional path.
		The optional line 0 serves to prevent a cascade of failure paths
		for the sake of performance
		by allowing threads to wait until transacting is safe again.}
	\label{fig:htm-fixed}
\end{figure}

\subsection{Modeling Transaction Failure}

Left unstated in interleavings such as $\ti@1-2c, \tj@1-3, \ti@4-7$%
\footnote{For the sake of example I reorder \ti's write to {\tt x} before \tj's part here.}
are HTM's execution semantics, namely:
\begin{enumerate}
	\item any modifications to shared state (such as $2c$) by \ti are not visible to \tj during its execution,
		despite \tj being executed afterwards, and
	\item all local and global state changes by \ti between lines $1$ and $2c$ are discarded when jumping to line $4$.
\end{enumerate}

\subsection{Modeling Transaction Atomicity}

@@

\section{Limitations}
\label{sec:warpzone}

{\bf Relaxed memory orderings.}
Our formalization of thread interleavings in Section~\ref{sec:design} does not account for read/write reorderings
possible on relaxed memory consistency architectures \cite{memory-consistency-models}.
In fact,
even after \cite{htm-mario}'s proposed fix,
our running example program is still incorrect on Total Store Ordering (TSO) architectures such as x86.
Despite stores being totally-ordered, x86 may still reorder stores after subsequent loads.
Accordingly, an execution of $8,9a,9b,9c$ in Figure~\ref{fig:htm-fixed}
may be locally visible to another thread as $9a,8,9b,9c$,
and hence an apparent interleaving of
\[
	\ti@1,\tj@1-5,\ti@7,\mathbf{\ti@9a},\tk@1-5,\mathbf{\ti@8},\ti@9b-B
\]
is possible
(reordered accesses highlighted in bold).
An {\tt mfence} barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}.
On Partial Store Ordering (PSO) architectures, even more barriers may be necessary.

Because Landslide's concurrency model includes only instruction-level thread nondeterminism,
not per-CPU memory buffer reorderings,
our current HTM implementation in Landslide cannot find this bug.
In fact, it erroneously verifies the corresponding test {\tt htm2} in 7 CPU-minutes,
with 294 distinct interleavings in total.
Recent work extended DPOR to incorporate TSO and PSO memory nondeterminism \cite{tsopso};
if incorporated into Landslide, we could find or verify the absence of such bugs.

% TODO should this come 1st or 2nd
{\bf Transaction failure codes.}
